\textbf{ Interpretable Gaussian Processes for Stellar Variability\\ I: Static Surfaces }

### 2020-12-15 tex/ms.tex
more unit tests
more unit tests, fixed eqn typos
more proofs
fix size integral
more figures

\features{open-source figures \codeicon; equation unit tests \testpassicon}

@@

@@

\subsection{Ensemble Analyses}
\label{sec:ensemble}

@@

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure a
            stellar light curve at a single random inclination (thin blue curves),
            at 3 random inclinations (thin orange curves), and at 10 random
            inclinations (thin green curves). The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing the star from many vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes.
            While this thought experiment is not possible in practice for
            a single star, it is extremely relevant to ensemble analyses of statistically
            similar stars.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

@@

\subsection{With limb darkening}
%
\label{sec:ld}
We must adjust our expression for the flux in the presence of limb darkening.
For any polynomial limb darkening law of the form
%
\begin{align}
    \label{eq:ld:I}
    \frac{I(\mu)}{I(\mu = 1)} = 1 - \sum_{n=1}^{n_\mathrm{max}} u_n(1 - \mu)^n
    \quad,
\end{align}
%
where $I$ is the intensity on the stellar surface,
$\mu = z = \sqrt{1 - x^2 - y^2}$ is the radial coordinate on the
projected disk, and $u_n$ is a limb darkening coefficient, the effect of
limb darkening on the stellar map can be expressed exactly as a linear
operation on the spherical harmonic coefficient vector
\citep{Luger2019}.
This includes the popular linear and quadractic limb darkening laws
and generalizes to \emph{any} limb darkening law in the limit
$n_\mathrm{max} \rightarrow \infty$. The linearity of the problem
can be understood by noting that all terms
in Equation~(\ref{eq:ld:I}) are strictly polynomials in $x$, $y$, and $z$,
all of which can be expressed exactly as sums of spherical harmonics
\citep{Luger2019}. When weighting the surface intensity by the limb darkening
profile, the resulting intensity is simply a product of spherical harmonics,
which is itself a linear combination of spherical harmonics.
Thus, given a limb darkening law
of degree $n_\mathrm{max}$ with coefficients $\mathbf{u}$,
we can construct a matrix $\mathbf{L}(\mathbf{u})$
that transforms a spherical harmonic vector $\mathbf{y}$ of degree
$l_\mathrm{max}$ to a limb-darkened spherical harmonic vector $\mathbf{y'}$
of degree $l_\mathrm{max} + n_\mathrm{max}$. As an example, consider
a map of degree $l_\mathrm{max} = 1$ and the linear limb darkening law
($n_\mathrm{max} = 1$) with coefficient vector $\mathbf{u} = ( u_1 )$.
The transformation matrix from $\mathbf{y}$ to $\mathbf{y'}$ is
%
\begin{proof}{}
    \label{eq:ld:L}
    \setstackgap{L}{1.5\baselineskip}
    \mathbf{L}(\mathbf{u}) =
    \frac{1}{1 - \frac{u_1}{3}}
    \begin{pmatrix}
        \quadquad1-u_1\quadquad & 0                       & \frac{u_1}{\sqrt{3}}    & 0                       \\
        0                       & \quadquad1-u_1\quadquad & 0                       & 0                       \\
        \frac{u_1}{\sqrt{3}}    & 0                       & \quadquad1-u_1\quadquad & 0                       \\
        0                       & 0                       & 0                       & \quadquad1-u_1\quadquad \\
        0                       & 0                       & 0                       & 0                       \\
        0                       & \frac{u_1}{\sqrt{5}}    & 0                       & 0                       \\
        0                       & 0                       & \frac{2u_1}{\sqrt{15}}  & 0                       \\
        0                       & 0                       & 0                       & \frac{u_1}{\sqrt{5}}    \\
        0                       & 0                       & 0                       & 0
    \end{pmatrix}
    \quad.
\end{proof}
%
The columns of $\mathbf{L}$ are constructed from the coefficient vectors of
each transformed spherical harmonic, which are in turn computed by
multiplying each spherical harmonic by the spherical harmonic representation
of the particular limb darkening law.

@@

\subsubsection{Rotation matrices}
\label{sec:wigner}
%
The Wigner rotation matrix for real spherical harmonics up to degree $l_{\mathrm{max}}$
may be written as the block-diagonal matrix
%
\begin{align}
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \mathbf{R} =
    \parenMatrixstack{
    \quad\quad \, \mathbf{R}^0 \, \quad\quad
     &              &              &        &                               \\
     & \mathbf{R}^1 &              &        &                               \\
     &              & \mathbf{R}^2 &        &                               \\
     &              &              & \ddots &                               \\
     &              &              &        & \mathbf{R}^{l_{\mathrm{max}}}
    }\quad,
\end{align}
%
where
%
\begin{align}
    \mathbf{R}^l = {\mathbf{U}^l}^{-1} \mathbf{D}^l \mathbf{U}^l
\end{align}
%
is the Wigner rotation matrix for a single spherical harmonic degree,
%
\begin{align}
    \label{eq:U}
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \mathbf{U}^l =
    \frac{1}{\sqrt{2}}
    \parenMatrixstack{
        \quad\quad\, \ddots \, \quad\quad\quad\quad
           &       &        &       &          &    &   &    & \Ddots \\
           & \imag &        &       &          &    &   & 1  &        \\
           &       & \imag  &       &          &    & 1 &    &        \\
           &       &        & \imag &          & 1  &   &    &        \\
           &       &        &       & \sqrt{2} &    &   &    &        \\
           &       &        & \imag &          & -1 &   &    &        \\
           &       & -\imag &       &          &    & 1 &    &        \\
           & \imag &        &       &          &    &   & -1 &        \\
    \Ddots &       &        &       &          &    &   &    & \ddots
    }
\end{align}
%
describes the transformation from complex to real spherical harmonics,
and $\mathbf{D}$ is the Wigner matrix for complex spherical harmonics,
whose terms are given by the expression
%
\begin{align}
    D_{m,m'}^l(\alpha, \beta, \gamma) = \mathrm{e}^{-\imag m' \alpha}
    d_{m,m'}^l(\beta) \mathrm{e}^{-\imag m \gamma}
\end{align}
%
where $\alpha$, $\beta$, and $\gamma$ are the Euler angles describing
the rotation in the
$\hat{\mathbf{z}}{-}\hat{\mathbf{y}}{-}\hat{\mathbf{z}}$ convention
and $\imag$ is the imaginary unit. The terms of the $d$-matrix
depend on powers of $\sin\left(\nicefrac{\beta}{2}\right)$
and $\cos\left(\nicefrac{\beta}{2}\right)$
\citep[c.f. Equation C15 in][]{Luger2019}, but it is convenient to use
the half-angle formula to express these terms instead as
%
\begin{proof}{test_dlmmp}
    d_{m,m'}^l(\beta) =
    \sum\limits_{i=0}^{2l} c_{m,m',i}^{l}
    \mathrm{sgn}(\sin\beta)^{i}
    (1 - \cos\beta)^{\frac{2l - i}{2}}
    (1 + \cos\beta)^\frac{i}{2}
\end{proof}
%
where
%
\begin{proof}{test_dlmmp}
    c_{m,m',i}^{l} & =
    \resizebox{.75\hsize}{!}{$
            \begin{cases}
                \dfrac{
                \left(-1\right)^{\frac{2l + 3m + m' - i}{2}}
                \sqrt{\left(l - m\right)!
                    \left(l + m\right)!
                    \left(l - m'\right)!
                    \left(l + m'\right)!}
                }{
                2^l
                \left(\frac{i - m - m'}{2}\right)!
                \left(\frac{i + m + m'}{2}\right)!
                \left(\frac{2l - i - m + m'}{2}\right)!
                \left(\frac{2l - i + m - m'}{2}\right)!
                }
                %
                 & m - m' - i \,\, \mathrm{even}
                %
                \\[0.5em]
                0
                %
                 & m - m' - i \,\, \mathrm{odd}
            \end{cases}
        $}
\end{proof}
%

@@

\begin{proof}{}

@@

    {({D^{l'}_{\hat{\mathbf{x}}}})_{}}_{{\mu'},m'}(\phi) \,

@@

        \mathrm{e}^{\frac{\imag \pi}{2}(m - \mu + {\mu'} - m')}

@@

        c_{{\mu'},m',i'}^{l'}

@@

To compute $\mu_\phi$, we differentiate Equation~(\ref{eq:phi-pdf}) with respect
to $\phi$, set the expression equal to zero, and solve
for $\phi$ to obtain
%
\begin{proof}{}
    \mu_\phi &= 2 \tan^{-1}
    \left(
    \sqrt{
        2 \alpha + \beta - 2 -
        \sqrt{
            4 \alpha^2
            - 8 \alpha
            - 6 \beta
            + 4 \alpha \beta
            + \beta^2
            + 5
        }
    }
    \right)
\end{proof}
%
where $\alpha$ and $\beta$ may be computed from $a$ and $b$ via
Equation~(\ref{eq:beta2gauss}).
%
To compute $\sigma_\phi$, we note that the variance of a Gaussian distribution
$\varphi(\phi)$
is the negative reciprocal of its curvature in log space:
%
\begin{align}
    \sigma^2 = -\left(\frac{\mathrm{d}^2 \ln \varphi(\phi)}{\mathrm{d}\phi^2}\right)^{-1}
\end{align}
%
We therefore twice differentiate the log of Equation~(\ref{eq:phi-pdf}), negate it,
take the reciprocal, and evaluate it at $\phi = \mu_\phi$ to obtain a local
approximation to the standard deviation of the distribution at the mode:
%
\begin{proof}{}
    \sigma_\phi &= \frac{\sin\mu_\phi}{
        \sqrt{1
            - \alpha
            + \beta
            + \left(\beta - 1\right) \cos\mu_\phi
            + \dfrac{\alpha - 1}{\cos\mu_\phi^2}}}
    \quad.
\end{proof}
%
For completeness, the inverse transform also has a closed form:
%
\begin{proof}{}
    \alpha &= \frac{2 + 4 \sigma_\phi^2 + (3 + 8 \sigma_\phi^2) \cos\mu_\phi + 2 \cos\left( 2\mu_\phi \right) + \cos\left( 3\mu_\phi \right)}{16 \sigma_\phi^2 \cos\left( \frac{\mu_\phi}{2} \right)^4} \\
    \beta &= \frac{\cos\mu_\phi + 2 \sigma_\phi^2 (3 + \cos\left( 2\mu_\phi \right)) - \cos\left( 3\mu_\phi \right)}{16 \sigma_\phi^2 \cos\left( \frac{\mu_\phi}{2} \right)^4}
    \quad.
\end{proof}

These transformations allow one to easily transform constraints on $a$ and
$b$ to constraints on the mode and approximate variance of the distribution
of spot latitudes. Importantly, they also allow us to compute the Jacobian
of the transform, given by
%
\begin{proof}{}
    J &=
    \frac{\partial{\mu_\phi}}{\partial{a}}
    \frac{\partial{\sigma_\phi}}{\partial{b}} -
    \frac{\partial{\mu_\phi}}{\partial{b}}
    \frac{\partial{\sigma_\phi}}{\partial{a}}
    %
    \\
    %
    &=\resizebox{.925\hsize}{!}{$
            \frac{
                \alpha \beta \left(1 + \cos\mu_\phi\right)^3
                \sin\left(2\mu_\phi\right)^3
            }{
                \sigma_\phi
                \left(
                2 \alpha + \beta - 3 +
                \left(2 \alpha + \beta - 1\right) \cos\mu_\phi
                \right)
                \left(
                2 \left( \alpha + \beta - 1\right) +
                3 (\beta - 1) \cos\mu_\phi
                - 2 \left( \alpha - \beta - 1 \right) \cos\left( 2\mu_\phi \right)
                + \left( \beta - 1\right) \cos\left( 3\mu_\phi \right)
                \right)^2
            }
        $}
    \nonumber
\end{proof}
%
While one could sample in $\mu_\phi$ and $\sigma_\phi$ directly, we instead
recommend sampling in the quantities $a$ and $b$ under a uniform prior on
$(0, 1)$ for both of them, then correcting for the implied prior on
$\mu_\phi$ and $\sigma_\phi$ by adding the log of the absolute value of $J$ to the log
likelihood. This corrects for the \emph{ad hoc}
prior on the latitude parameters introduced by our particular choice of
parametrization, enforcing instead a uniform prior on the quantities $\mu_\phi$
and $\sigma_\phi$.

@@

\begin{proof}{}

@@

    &\equiv \mathbf{e}_{y'}

@@

\begin{proof}{}

@@

@@

    (-1)^{i + 1}

@@

    \frac{(-1)^{i+1}}{l-\frac{i}{2}+1}

@@

\subsection{Second moment}
\label{sec:inc-mom2}
%
The expression for the second moment is
%
\begin{proof}{}
    \mathbf{E}_I
    & \equiv
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathbf{E}_y \,
    \pmb{\mathcal{A}}^\top(I, \bar{\pmb{\theta}}_\star) \,
    p(I) \,
    \mathrm{d}I
\end{proof}
%
where
%
\begin{proof}{}
    \mathbf{E}_y
    & \equiv
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
\end{proof}
%
is the second moment of the GP in the spherical harmonic basis
(Equation~\ref{eq:exp_yy}). We can use Equations~(\ref{eq:Arows})
and (\ref{eq:akT})
to express the element at index $k, k'$ (corresponding to the covariance of the
GP between times $t = t_k$ and $t' = t_{k'}$) as
%
\begin{align}
    \label{eq:Eikkp}
    \left(E_I\right)_{k, k'}
     & =
    \int
    \mathbf{a}_k^\top
    \mathbf{E}_y \,
    \mathbf{a}_{k'}
    p(I) \,
    \mathrm{d}I
    \nonumber \\
     & =
    \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \left(\mathbf{E}_{y''}\right)_{k,k'}
    \mathbf{A_1}^\top \,
    \mathbf{r}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:Eyppkkp}
    \left(\mathbf{E}_{y''}\right)_{k,k'}
     & =
    \int
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
    \left(\mathbf{E}_{y'}\right)_{k,k'} \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(-I\right) \,
    p(I) \,
    \mathrm{d}I
\end{align}
%
is the expectation of $\mathbf{y}\,\mathbf{y}^\top$ in the
observer's frame at times $t = t_k$ and $t' = t_{k'}$
and
%
\begin{align}
    \left(\mathbf{E}_{y'}\right)_{k,k'}
     & \equiv
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{E}_y \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(\frac{\pi}{2}\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}^\top\left(\frac{2\pi}{P}t_{k'}\right)
\end{align}
%
is the expectation of $\mathbf{y}\,\mathbf{y}^\top$ in the
polar frame at times $t = t_k$ and $t' = t_{k'}$.
%
The rest of the computation follows what we did in \S\ref{sec:lat},
except that the number of operations required to compute $\mathbf{E}_I$
is a factor of $K^2$ larger than in the computation of expectations
like $\mathbf{E}_\phi$ (Equation~\ref{eq:Ephi}). That is because we
must compute the integral of all terms of a matrix
for \emph{each} of the $K^2$ elements of $\mathbf{E}_I$.
We discuss in \S\ref{sec:inc-speedup} below strategies that can drastically
improve the computational scaling of marginalizing over the inclination.

@@

\begin{proof}{}

@@

    (1 - \cos\phi)^{l + l' - \frac{i+i'}{2}}

@@

    (-1)^{i+i'+1}

@@

    \frac{(-1)^{i+i'+1}}{l+l'-\frac{i+i'}{2}+1}

>>>>

\features{open-source figures \codeicon; equation unit tests: \input{tests/tally}}

@@

\xxx{Brief paper outline here.}

@@

\subsection{Ensemble analyses}
\label{sec:ensemble}

@@

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble.pdf}
        \oscaption{nullspace_ensemble}{%
            Posterior shrinkage as a function of spherical harmonic degree
            for a hypothetical scenario in which an observer can measure a
            stellar light curve at 1 (thin blue curves),
            3 (thin orange curves), 10 (thin green curves), and 30 (thin red
            curves) random inclinations. The thicker curves correspond to
            the mean shrinkage for each experiment.
            The information content in the light curve of a star observed
            from a single vantage point approaches zero as $l$
            increases. However, observing the star from many vantage points
            allows one to recover nearly all of the information in the
            even spherical harmonic modes.
            While this thought experiment is not possible in practice for
            a single star, it is extremely relevant to ensemble analyses of statistically
            similar stars.
            \label{fig:nullspace_ensemble}
        }
    \end{centering}
\end{figure}

@@

\subsection{Limb darkening}

See Figures~\ref{fig:nullspace_ensemble_ld} and \ref{fig:nullspace_ensemble_ld_var}.

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld.pdf}
        \oscaption{nullspace_ensemble_ld}{%
            Same as Figure~\ref{fig:nullspace_ensemble}, but for limb-darkened
            stars with coefficients $u_1 = 0.5$ and $u_2 = 0.25$.
            \label{fig:nullspace_ensemble_ld}
        }
    \end{centering}
\end{figure}

\begin{figure}[ht!]
    \begin{centering}
        \includegraphics[width=\linewidth]{figures/nullspace_ensemble_ld_var.pdf}
        \oscaption{nullspace_ensemble_ld_var}{%
            Same as Figure~\ref{fig:nullspace_ensemble_ld}, but allowing for
            a 10\% variation in each of the two limb darkening coefficients
            across different observations.
            \label{fig:nullspace_ensemble_ld_var}
        }
    \end{centering}
\end{figure}

@@

\subsection{With limb darkening}
% http://dx.doi.org/10.1051/0004-6361/200913675
%
\label{sec:ld}
We must adjust our expression for the flux in the presence of limb darkening.
For any polynomial limb darkening law of the form
%
\begin{align}
    \label{eq:ld:I}
    \frac{I(\mu)}{I(\mu = 1)} = 1 - \sum_{n=1}^{n_\mathrm{max}} u_n(1 - \mu)^n
    \quad,
\end{align}
%
where $I$ is the intensity on the stellar surface,
$\mu = z = \sqrt{1 - x^2 - y^2}$ is the radial coordinate on the
projected disk, and $u_n$ is a limb darkening coefficient, the effect of
limb darkening on the stellar map can be expressed exactly as a linear
operation on the spherical harmonic coefficient vector
\citep{Luger2019}.
This includes the popular linear and quadractic limb darkening laws
and generalizes to \emph{any} limb darkening law in the limit
$n_\mathrm{max} \rightarrow \infty$. The linearity of the problem
can be understood by noting that all terms
in Equation~(\ref{eq:ld:I}) are strictly polynomials in $x$, $y$, and $z$,
all of which can be expressed exactly as sums of spherical harmonics
\citep{Luger2019}. When weighting the surface intensity by the limb darkening
profile, the resulting intensity is simply a product of spherical harmonics,
which is itself a linear combination of spherical harmonics.
Thus, given a limb darkening law
of degree $n_\mathrm{max}$ with coefficients $\mathbf{u}$,
we can construct a matrix $\mathbf{L}(\mathbf{u})$
that transforms a spherical harmonic vector $\mathbf{y}$ of degree
$l_\mathrm{max}$ to a limb-darkened spherical harmonic vector $\mathbf{y'}$
of degree $l_\mathrm{max} + n_\mathrm{max}$. As an example, consider
a map of degree $l_\mathrm{max} = 1$ and the linear limb darkening law
($n_\mathrm{max} = 1$) with coefficient vector $\mathbf{u} = ( u_1 )$.
The transformation matrix from $\mathbf{y}$ to $\mathbf{y'}$ is
%
\begin{proof}{}
    \label{eq:ld:L}
    \setstackgap{L}{1.5\baselineskip}
    \mathbf{L}(\mathbf{u}) =
    \frac{1}{1 - \frac{u_1}{3}}
    \begin{pmatrix}
        \quadquad1-u_1\quadquad & 0                       & \frac{u_1}{\sqrt{3}}    & 0                       \\
        0                       & \quadquad1-u_1\quadquad & 0                       & 0                       \\
        \frac{u_1}{\sqrt{3}}    & 0                       & \quadquad1-u_1\quadquad & 0                       \\
        0                       & 0                       & 0                       & \quadquad1-u_1\quadquad \\
        0                       & 0                       & 0                       & 0                       \\
        0                       & \frac{u_1}{\sqrt{5}}    & 0                       & 0                       \\
        0                       & 0                       & \frac{2u_1}{\sqrt{15}}  & 0                       \\
        0                       & 0                       & 0                       & \frac{u_1}{\sqrt{5}}    \\
        0                       & 0                       & 0                       & 0
    \end{pmatrix}
    \quad.
\end{proof}
%
The columns of $\mathbf{L}$ are constructed from the coefficient vectors of
each transformed spherical harmonic, which are in turn computed by
multiplying each spherical harmonic by the spherical harmonic representation
of the particular limb darkening law.

@@

\subsubsection{Rotation matrices}
\label{sec:wigner}
%
The Wigner rotation matrix for real spherical harmonics up to degree $l_{\mathrm{max}}$
may be written as the block-diagonal matrix
%
\begin{align}
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \mathbf{R} =
    \parenMatrixstack{
    \quad\quad \, \mathbf{R}^0 \, \quad\quad
     &              &              &        &                               \\
     & \mathbf{R}^1 &              &        &                               \\
     &              & \mathbf{R}^2 &        &                               \\
     &              &              & \ddots &                               \\
     &              &              &        & \mathbf{R}^{l_{\mathrm{max}}}
    }\quad,
\end{align}
%
where
%
\begin{align}
    \mathbf{R}^l = {\mathbf{U}^l}^{-1} \mathbf{D}^l \mathbf{U}^l
\end{align}
%
is the Wigner rotation matrix for a single spherical harmonic degree,
%
\begin{align}
    \label{eq:U}
    \setstackgap{L}{1.25\baselineskip}
    \fixTABwidth{T}
    \mathbf{U}^l =
    \frac{1}{\sqrt{2}}
    \parenMatrixstack{
        \quad\quad\, \ddots \, \quad\quad\quad\quad
           &       &        &       &          &    &   &    & \Ddots \\
           & \imag &        &       &          &    &   & 1  &        \\
           &       & \imag  &       &          &    & 1 &    &        \\
           &       &        & \imag &          & 1  &   &    &        \\
           &       &        &       & \sqrt{2} &    &   &    &        \\
           &       &        & \imag &          & -1 &   &    &        \\
           &       & -\imag &       &          &    & 1 &    &        \\
           & \imag &        &       &          &    &   & -1 &        \\
    \Ddots &       &        &       &          &    &   &    & \ddots
    }
\end{align}
%
describes the transformation from complex to real spherical harmonics,
and $\mathbf{D}$ is the Wigner matrix for complex spherical harmonics,
whose terms are given by the expression
%
\begin{align}
    D_{m,m'}^l(\alpha, \beta, \gamma) = \mathrm{e}^{-\imag m' \alpha}
    d_{m,m'}^l(\beta) \mathrm{e}^{-\imag m \gamma}
\end{align}
%
where $\alpha$, $\beta$, and $\gamma$ are the Euler angles describing
the rotation in the
$\hat{\mathbf{z}}{-}\hat{\mathbf{y}}{-}\hat{\mathbf{z}}$ convention
and $\imag$ is the imaginary unit. The terms of the $d$-matrix
depend on powers of $\sin\left(\nicefrac{\beta}{2}\right)$
and $\cos\left(\nicefrac{\beta}{2}\right)$
\citep[c.f. Equation C15 in][]{Luger2019}, but it is convenient to use
the half-angle formula to express these terms instead as
%
\begin{proof}{test_dlmmp}
    d_{m,m'}^l(\beta) =
    \sum\limits_{i=0}^{2l} c_{m,m',i}^{l}
    \mathrm{sgn}(\sin\beta)^{i}
    (1 - \cos\beta)^{\frac{2l - i}{2}}
    (1 + \cos\beta)^\frac{i}{2}
\end{proof}
%
where
%
\begin{proof}{test_dlmmp}
    c_{m,m',i}^{l} & =
    \resizebox{.75\hsize}{!}{$
            \begin{cases}
                \dfrac{
                \left(-1\right)^{\frac{2l - m + m' - i}{2}}
                \sqrt{\left(l - m\right)!
                    \left(l + m\right)!
                    \left(l - m'\right)!
                    \left(l + m'\right)!}
                }{
                2^l
                \left(\frac{i - m - m'}{2}\right)!
                \left(\frac{i + m + m'}{2}\right)!
                \left(\frac{2l - i - m + m'}{2}\right)!
                \left(\frac{2l - i + m - m'}{2}\right)!
                }
                %
                 & m - m' - i \,\, \mathrm{even}
                %
                \\[0.5em]
                0
                %
                 & m - m' - i \,\, \mathrm{odd}
            \end{cases}
        $}
\end{proof}
%

@@

\begin{proof}{test_bigEphi}

@@

    {({D^{l'}_{\hat{\mathbf{x}}}})_{}}_{m',{\mu'}}(\phi) \,

@@

        \mathrm{e}^{\frac{\imag \pi}{2}(m - \mu + m' - {\mu'})}

@@

        c_{m',{\mu'},i'}^{l'}

@@

To compute $\mu_\phi$, we differentiate Equation~(\ref{eq:phi-pdf}) with respect
to $\phi$, set the expression equal to zero, and solve
for $\phi$ to obtain
%
\begin{proof}{test_lat_transforms}
    \mu_\phi &= 2 \tan^{-1}
    \left(
    \sqrt{
        2 \alpha + \beta - 2 -
        \sqrt{
            4 \alpha^2
            - 8 \alpha
            - 6 \beta
            + 4 \alpha \beta
            + \beta^2
            + 5
        }
    }
    \right)
\end{proof}
%
where $\alpha$ and $\beta$ may be computed from $a$ and $b$ via
Equation~(\ref{eq:beta2gauss}).
%
To compute $\sigma_\phi$, we note that the variance of a Gaussian distribution
$\varphi(\phi)$
is the negative reciprocal of its curvature in log space:
%
\begin{align}
    \sigma^2 = -\left(\frac{\mathrm{d}^2 \ln \varphi(\phi)}{\mathrm{d}\phi^2}\right)^{-1}
\end{align}
%
We therefore twice differentiate the log of Equation~(\ref{eq:phi-pdf}), negate it,
take the reciprocal, and evaluate it at $\phi = \mu_\phi$ to obtain a local
approximation to the standard deviation of the distribution at the mode:
%
\begin{proof}{test_lat_transforms}
    \sigma_\phi &= \frac{\sin\mu_\phi}{
        \sqrt{1
            - \alpha
            + \beta
            + \left(\beta - 1\right) \cos\mu_\phi
            + \dfrac{\alpha - 1}{\cos\mu_\phi^2}}}
    \quad.
\end{proof}
%
For completeness, the inverse transform also has a closed form:
%
\begin{proof}{test_lat_transforms}
    \alpha &= \frac{2 + 4 \sigma_\phi^2 + (3 + 8 \sigma_\phi^2) \cos\mu_\phi + 2 \cos\left( 2\mu_\phi \right) + \cos\left( 3\mu_\phi \right)}{16 \sigma_\phi^2 \cos\left( \frac{\mu_\phi}{2} \right)^4} \\
    \beta &= \frac{\cos\mu_\phi + 2 \sigma_\phi^2 (3 + \cos\left( 2\mu_\phi \right)) - \cos\left( 3\mu_\phi \right)}{16 \sigma_\phi^2 \cos\left( \frac{\mu_\phi}{2} \right)^4}
    \quad.
\end{proof}

These transformations allow one to easily transform constraints on $a$ and
$b$ to constraints on the mode and approximate variance of the distribution
of spot latitudes. Importantly, they also allow us to compute the Jacobian
of the transform, given by
%
\begin{proof}{test_lat_jac}
    J &=
    \frac{\partial{\mu_\phi}}{\partial{a}}
    \frac{\partial{\sigma_\phi}}{\partial{b}} -
    \frac{\partial{\mu_\phi}}{\partial{b}}
    \frac{\partial{\sigma_\phi}}{\partial{a}}
    %
    \\
    %
    &=\resizebox{.925\hsize}{!}{$
            \frac{
                \left(K_{10} - K_{00}\right) \left(K_{11} - K_{10}\right)
                \alpha \beta \left(1 + \cos\mu_\phi\right)^3
                \sin\left(2\mu_\phi\right)^3
            }{
                \sigma_\phi
                \left(
                2 \alpha + \beta - 3 +
                \left(2 \alpha + \beta - 1\right) \cos\mu_\phi
                \right)
                \left(
                2 \left( \alpha + \beta - 1\right) +
                3 (\beta - 1) \cos\mu_\phi
                - 2 \left( \alpha - \beta - 1 \right) \cos\left( 2\mu_\phi \right)
                + \left( \beta - 1\right) \cos\left( 3\mu_\phi \right)
                \right)^2
            }
        $}
    \nonumber
\end{proof}
%
While one could sample in $\mu_\phi$ and $\sigma_\phi$ directly, we instead
recommend sampling in the quantities $a$ and $b$ under a uniform prior on
$(0, 1)$ for both of them, then correcting for the implied prior on
$\mu_\phi$ and $\sigma_\phi$ by adding the log of the absolute value of $J$ to the log
likelihood. This corrects for the \emph{ad hoc}
prior on the latitude parameters introduced by our particular choice of
parametrization, enforcing instead a uniform prior on the quantities $\mu_\phi$
and $\sigma_\phi$.

@@

@@

     & \equiv \mathbf{e}_{y'}

@@

\begin{proof}{test_eI}

@@

\begin{proof}{test_eI}

@@

@@

    \frac{(-1)^{i}}{l-\frac{i}{2}+1}

@@

\subsection{Second moment}
\label{sec:inc-mom2}
%
The expression for the second moment is
%
\begin{align}
    \mathbf{E}_I
     & \equiv
    \int
    \pmb{\mathcal{A}}(I, \bar{\pmb{\theta}}_\star) \,
    \mathbf{E}_y \,
    \pmb{\mathcal{A}}^\top(I, \bar{\pmb{\theta}}_\star) \,
    p(I) \,
    \mathrm{d}I
\end{align}
%
where
%
\begin{align}
    \mathbf{E}_y
     & \equiv
    \mathrm{E} \Big[ \mathbf{y} \, \mathbf{y}^\top \, \Big| \, \pmb{\theta}_\bullet \Big]
\end{align}
%
is the second moment of the GP in the spherical harmonic basis
(Equation~\ref{eq:exp_yy}). We can use Equations~(\ref{eq:Arows})
and (\ref{eq:akT})
to express the element at index $k, k'$ (corresponding to the covariance of the
GP between times $t = t_k$ and $t' = t_{k'}$) as
%
\begin{align}
    \label{eq:Eikkp}
    \left(E_I\right)_{k, k'}
     & =
    \int
    \mathbf{a}_k^\top
    \mathbf{E}_y \,
    \mathbf{a}_{k'}
    p(I) \,
    \mathrm{d}I
    \nonumber \\
     & =
    \mathbf{r}^\top \,
    \mathbf{A_1} \,
    \left(\mathbf{E}_{y''}\right)_{k,k'}
    \mathbf{A_1}^\top \,
    \mathbf{r}
    \quad,
\end{align}
%
where
%
\begin{align}
    \label{eq:Eyppkkp}
    \left(\mathbf{E}_{y''}\right)_{k,k'}
     & =
    \int
    \mathbf{R}_{\hat{\mathbf{x}}}\left(-I\right) \,
    \left(\mathbf{E}_{y'}\right)_{k,k'} \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(-I\right) \,
    p(I) \,
    \mathrm{d}I
\end{align}
%
is the expectation of $\mathbf{y}\,\mathbf{y}^\top$ in the
observer's frame at times $t = t_k$ and $t' = t_{k'}$
and
%
\begin{align}
    \left(\mathbf{E}_{y'}\right)_{k,k'}
     & \equiv
    \mathbf{R}_{\hat{\mathbf{z}}}\left(\frac{2\pi}{P}t_k\right) \,
    \mathbf{R}_{\hat{\mathbf{x}}}\left(\frac{\pi}{2}\right) \,
    \mathbf{E}_y \,
    \mathbf{R}_{\hat{\mathbf{x}}}^\top\left(\frac{\pi}{2}\right) \,
    \mathbf{R}_{\hat{\mathbf{z}}}^\top\left(\frac{2\pi}{P}t_{k'}\right)
\end{align}
%
is the expectation of $\mathbf{y}\,\mathbf{y}^\top$ in the
polar frame at times $t = t_k$ and $t' = t_{k'}$.
%
The rest of the computation follows what we did in \S\ref{sec:lat},
except that the number of operations required to compute $\mathbf{E}_I$
is a factor of $K^2$ larger than in the computation of expectations
like $\mathbf{E}_\phi$ (Equation~\ref{eq:Ephi}). That is because we
must compute the integral of all terms of a matrix
for \emph{each} of the $K^2$ elements of $\mathbf{E}_I$.
We discuss in \S\ref{sec:inc-speedup} below strategies that can drastically
improve the computational scaling of marginalizing over the inclination.

@@

\begin{proof}{test_bigEI}

@@

    (1 - \cos I)^{l + l' - \frac{i+i'}{2}}
    (1 + \cos I)^\frac{i+i'}{2}

@@

    \int_{0}^{1}

@@

    \frac{(-1)^{i+i'}}{l+l'-\frac{i+i'}{2}+1}
